---
title: "MM923 Data Analytics in R - Part 1: Data Management"
author: "Nathan Ormond"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: united
    highlight: tango
    code_folding: show
    df_print: paged
---

# MM923 Data Analytics in R - Part 1: Data Management
This notebook contains code and documentation for cleaning and preparing the weather data for analysis. It addresses the first part of the assignment (20 marks).


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Set working directory to project root using relative path
knitr::opts_knit$set(root.dir = here::here())
```

# Part 1: Data Management (20 marks)

This notebook implements the data cleaning and preparation steps required for Part 1 of the MM923 assessment.

## Packages

```{r load-packages}
# Tidyverse for data manipulation
library(tidyverse)  
# Seed for reproducibility
set.seed(123)
```

## Raw Data

```{r load-data}
load("data/raw/forecasts.RData")
load("data/raw/cities.RData")
head(forecasts)
head(cities)
```

Examining the structure of the data sets: 

```{r examine-structure}
str(forecasts)
str(cities)

summary(forecasts)
summary(cities)
```

## Cleaning Forecasts
From assignment:
1. Create unique variables for State and City
2. Make each level of Measurement have its own column with measure from Response
3. Obtain only rows with accurate observed temp
4. Retain only information for the year 2021
5. Calculate the average temperature of each city and store in a variable avg_temp rounded to 3 decimal places
6. Retain only one row from each city in each state

```{r clean-forecasts}
# Step 1: Create unique variables for State and City
forecasts_clean <- forecasts %>%
  separate(col = `State City`, into = c("State", "City"), sep = ":")

# Check the result of Step 1
head(forecasts_clean)

# Step 2: Make each Measurement have its own column
forecasts_clean <- forecasts_clean %>%
  pivot_wider(names_from = Measurement, values_from = Response)

# Check the result of Step 2
head(forecasts_clean)

# Step 3: Obtain only rows with accurate observed temp
forecasts_clean <- forecasts_clean %>%
  filter(`possible error` == "none") %>%
  filter(!is.na(`observed temp`))

# Check the result of Step 3
head(forecasts_clean)

# Step 4: Retain only information for the year 2021
forecasts_clean <- forecasts_clean %>%
  filter(format(date, "%Y") == "2021")

# Check the result of Step 4
head(forecasts_clean)
min(forecasts_clean$date)
max(forecasts_clean$date)

# Step 5: Calculate average temperature rounded to 3 decimal places
forecasts_clean <- forecasts_clean %>%
  group_by(State, City) %>%
  summarise(avg_temp = round(mean(`observed temp`, na.rm = TRUE), 3))

# Check the result of Step 5
head(forecasts_clean)

# Step 6: Retain only one row from each city in each state
# This is already done by the group_by and summarise operations
# But we'll explicitly check for duplicates
duplicated_rows <- forecasts_clean %>%
  group_by(State, City) %>%
  filter(n() > 1)

print(paste("Number of duplicated State-City combinations:", nrow(duplicated_rows)))
```

## Cleaning Cities 
Assignment:
1. Create a new variable called koppen which uses variable names from columns 10 to 25 as levels
2. Store the responses in a variable called avg_annual_precip, dropping rows with missing values
3. Combine with the tidy forecasts data retaining cities that appear in both datasets
4. Retain only specific variables (State, City, lon, lat, koppen, elevation, distance to coast, wind, elevation change four, elevation change eight, avg_annual_precip, avg_temp)
```{r clean-cities}
# Step 1-2: Create koppen variable and avg_annual_precip
cities_clean <- cities %>%
  # Identify the Köppen climate columns (10-25)
  pivot_longer(cols = names(cities)[10:25], 
               names_to = "koppen", 
               values_to = "avg_annual_precip") %>%
  # Drop rows with missing values in avg_annual_precip
  filter(!is.na(avg_annual_precip)) %>%
  # Create State and City variables matching forecasts
  rename(State = state, City = city)

# Check the result of Step 1-2
head(cities_clean)
```

## Combining Datasets
Now we'll combine both cleaned datasets and keep only cities that appear in both.

```{r combine-datasets}
# Step 3: Combine the datasets keeping only cities in both
weather_data <- inner_join(forecasts_clean, cities_clean, by = c("State", "City"))

# Step 4: Retain only the specified variables
weather_data <- weather_data %>%
  select(State, City, lon, lat, koppen, elevation, `distance to coast`, 
         wind, `elevation change four`, `elevation change eight`, 
         avg_annual_precip, avg_temp)

# Check the final dataset
head(weather_data)
dim(weather_data)
```

## Validating Against Reference Data

```{r validate-data}
# Load the reference data
load("data/raw/weather.RData")

# Check if our cleaned data matches the reference data
identical(weather_data, weather)

# If not identical, investigate differences
if(!identical(weather_data, weather)) {
  # Check column names
  cat("Column names match:", all.equal(names(weather_data), names(weather)), "\n")
  
  # Check dimensions
  cat("Dimensions match:", all.equal(dim(weather_data), dim(weather)), "\n")
  
  # If dimensions don't match, investigate which cities might be missing
  if(!identical(dim(weather_data), dim(weather))) {
    # Check difference in rows
    cat("Number of rows in our data:", nrow(weather_data), "\n")
    cat("Number of rows in reference data:", nrow(weather), "\n")
    
    # Compare city sets
    our_cities <- paste(weather_data$State, weather_data$City, sep="-")
    ref_cities <- paste(weather$State, weather$City, sep="-")
    
    cat("Cities in reference but not in our data:", "\n")
    print(setdiff(ref_cities, our_cities))
    
    cat("Cities in our data but not in reference:", "\n")
    print(setdiff(our_cities, ref_cities))
  }
  
  # Check a sample of values to see if there are numeric differences
  cat("First 5 rows of our data:", "\n")
  print(head(weather_data, 5))
  
  cat("First 5 rows of reference data:", "\n")
  print(head(weather, 5))
}
```

## Saving Clean Data
```{r save-data}
# Save the cleaned data
saveRDS(weather_data, "data/processed/weather_clean.rds")

# Also save as RData for consistency with the original format
weather <- weather_data  # Create an object with the expected name
save(weather, file = "data/processed/weather.RData")
```

## Summary

Completed
1. [x] Loaded the raw forecasts and cities data
2. [ ] Cleaned the forecasts data according to specifications
3. [ ] Cleaned the cities data according to specifications
4. [ ] Combined the datasets and retained only necessary variables
5. [ ] Validated our cleaned data against the reference data
6. [ ] Saved the cleaned data for use in subsequent analyses

The resulting dataset contains weather information for US cities, including average temperature, geographical information, and climate classifications.


